Efficient entropy stable Gauss collocation methods

### 2018-08-30 docs/ESDG_GQ/esdg_gq.tex

\newcommand{\Oh}{{\Omega_h}}
\renewcommand{\L}{L^2\LRp{\Omega}}
\newcommand{\LK}{L^2\LRp{D^k}}
\newcommand{\LdK}{L^2\LRp{\partial D^k}}
\newcommand{\Dhat}{\widehat{D}}
\newcommand{\Lhat}{L^2\LRp{\Dhat}}

@@

\newcommand{\LinfDk}{L^{\infty}\LRp{D^k}}
\newcommand{\diag}[1]{{\rm diag}\LRp{#1}}
\newcommand{\Ksub}{K_{\rm sub}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

@@

\begin{abstract}
The construction of high order entropy stable collocation schemes on quadrilateral and hexahedral elements has relied on the use of Gauss-Legendre-Lobatto collocation points \cite{fisher2013high, carpenter2014entropy, gassner2016split} and their equivalence with summation-by-parts (SBP) finite difference operators \cite{gassner2013skew}.  In this work, we show how to efficiently generalize the construction of semi-discretely entropy stable schemes on tensor product elements to Gauss points and generalized SBP operators.  Numerical experiments suggest that the use of Gauss points can significant improve accuracy on curved meshes.  
\end{abstract}

@@

%Time dependent nonlinear conservation laws are ubiquitous in computational fluid dynamics (CFD).  These conservation laws include the compressible Euler and Navier-Stokes equations, which govern compressible fluid behavior within aerospace applications, as well as the equations governing coastal flooding and geophysical flows (shallow water), and those describing astrophysical phenomena (magnetohydrodynamics).  The utility of these simulations depends both on the reliability and accuracy of numerical methods for the solution of the associated nonlinear partial differential equations (PDEs).  
Time dependent nonlinear conservation laws are ubiquitous in computational fluid dynamics (CFD), for which high order methods are increasingly of interest.  Such methods are more accurate per degree of freedom than low order methods, while also possessing much smaller numerical dispersion and dissipation errors.  This makes high order methods especially well suited to time-dependent simulations.  In this work, we focus specifically on discontinuous Galerkin methods on unstructured quadrilateral and hexahedral meshes.  These methods combine properties of high order approximations with the geometric flexibility of unstructured meshing.  %When paired with explicit time-stepping methods, high order DG and finite element methods are also well-suited for many-core and GPU computing \cite{klockner2009nodal, swirydowicz2017acceleration, wintermeyer2018entropy}.

@@

The instability of high order methods is rooted in the fact that discretizations of nonlinear conservation laws do not typically satisfy a discrete analogue of the conservation or dissipation of energy (entropy).  While the numerical dissipation present for low order methods serves as a stabilizing mechanism, the low of numerical dissipation in high order methods reveals the lack of an inherent statement of stability.  The dissipation of entropy serves as an energetic principle for nonlinear conservation laws \cite{dafermos2005compensated}, and requires the use of the chain rule in its proof.  The instability of high order methods is tied to the fact that, when discretizing systems of nonlinear PDEs, the chain rule does not typically hold at the discrete level.

The lack of a chain rule was circumvented by using a non-standard ``flux differencing'' formulation \cite{fisher2013high, carpenter2014entropy, gassner2016split, gassner2017br1}, which is key to constructing semi-discretely entropy stable high order schemes on unstructured quadrilateral and hexahedral meshes.  Flux differencing replaces the derivative of the nonlinear flux with the discrete differentiation of an auxiliary quantity.  This auxiliary quantity is computed through the evaluation of a two-point entropy conservative flux \cite{tadmor1987numerical} using pairs of solution values at quadrature points.  These entropy stable schemes were later extended to non-tensor product elements using GLL-like quadrature points on triangles and tetrahedra \cite{chen2017entropy, crean2018entropy}.  More recently, the construction of efficient entropy stable schemes was extended to more arbitrary choices of basis and quadrature \cite{chan2017discretely, chan2018discretely}.

@@

The main motivation for exploring tensor product (quadrilateral and hexahedral) elements is the significant reduction in the number of operations required compared to high order entropy stable schemes on simplicial meshes \cite{chan2017discretely, chan2018discretely}.  Entropy stability schemes on simplicial elements require evaluating two-point fluxes between solution states at all quadrature points on an element.  The number of quadrature points on a simplex scales as $O(N^d)$ in $d$ dimensions, and results in $O(N^{2d})$ two-point flux evaluations per element.  In contrast, entropy stable schemes on quadrilateral and hexahedral elements require only the evaluation of two-point fluxes along lines of nodes due to a tensor product structure, resulting in $O(N^{d+1})$ evaluations in $d$ dimensions.

@@

We are interested in methods for the numerical solution of systems of conservation laws in $d$ dimensions
\begin{equation}
\pd{\bm{u}}{t} + \sum_{i=1}^d \pd{\bm{f}_i\LRp{\bm{u}}}{x_i} = 0,
\label{eq:nonlinpde}
\end{equation}
where $\bm{u}$ denotes the conservative variables, $\bm{f}_i(\bm{u})$ are nonlinear fluxes, and $x_i$ denotes the $i$th coordinate.  Many physically motivated conservation laws admit a statement of stability involving a convex scalar entropy $S(\bm{u})$.  We first define the entropy variables $\bm{v}(\bm{u})$ to be the gradient of the entropy $S(\bm{u})$ with respect to the conservative variables 
\[
\bm{v} = \pd{S(\bm{u})}{\bm{u}}.  
\]
For a convex entropy, $\bm{v}(\bm{u})$ defines an invertible mapping from conservative to entropy variables.  We denote the inverse of this mapping (from entropy to conservative variables) by $\bm{u}(\bm{v})$.

At the continuous level, it can be shown (for example, in \cite{dafermos2005compensated}) that vanishing viscosity solutions to (\ref{eq:nonlinpde}) satisfy the strong form of an entropy inequality 
\begin{gather}
\pd{S(\bm{u})}{t} + \sum_{i=1}^d \pd{F_i(\bm{u})}{x_i} \leq 0,
\label{eq:entropyineq}
\end{gather}
where $F_i$ denotes the $i$th scalar entropy flux function and $\psi_i(\bm{u})$ denotes the $i$th entropy potential
\[
F_i(\bm{u}) = \bm{v}^T\pd{\bm{f}_i}{x_i}, \qquad \psi_i(\bm{u}) = \bm{v}^T\bm{f}_i(\bm{u}) - F_i(\bm{u}). 
\]
Integrating (\ref{eq:entropyineq}) over a domain $\Omega$ and applying the divergence theorem yields an integrated entropy inequality
\begin{equation}
\int_{\Omega} \pd{S(\bm{u})}{t} + \int_{\partial \Omega} \sum_{i=1}^d \bm{n}_i \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})} \leq 0,
\label{eq:weakentropyineq}
\end{equation}
where $\partial \Omega$ denotes the boundary of $\Omega$ and $\bm{n}_i$ denotes the $i$th component of the outward normal on $\partial \Omega$.  Roughly speaking, this implies that the time rate of change of entropy is less than or equal to the flux of entropy through the boundary.

@@

This comparison is motivated by the accuracy of each respective quadrature rule.  While $(N+1)$-point GLL quadrature rules are exact for polynomial integrands of degree $(2N-1)$, $(N+1)$-point Gauss quadrature rules are exact for polynomials of degree $(2N+1)$.  This higher accuracy of Gauss quadrature has been shown to translate to lower errors and slightly improved rates of convergence in simulations of wave propagation and fluid flow \cite{kopriva2010quadrature, hindenlang2012explicit, chan2015gpu}.  However, Gauss points have not been widely used to construct entropy stable discretizations due to the lack of efficient, stable, and high order accurate inter-element coupling terms, known as simultaneous approximation terms (SBP-SAT) in the finite difference literature \cite{fernandez2014review, crean2017high, fernandez2018simultaneous}.  SBP-SATs for Gauss points are non-compact, in the sense that they introduce all-to-all coupling between degrees of freedom on neighboring elements.  This results in greater communication between elements, as well as a significantly larger number of two-point flux evaluations and floating point operations.

It is possible to realize the improved accuracy of Gauss points while avoiding non-compact SBP-SATs through a staggered grid formulation, where the solution is stored at Gauss nodes but interpolated to a set of higher degree $(N+2)$ GLL ``flux'' points for computation \cite{parsani2016entropy}.  Because GLL nodes include boundary points, compact and high order accurate SBP-SAT terms can be constructed for the flux points.  After performing computations on the flux points, the results are interpolated back to Gauss points and used to evolve the solution forward in time.  Figure~\ref{fig:nodesets} shows an illustration of GLL, staggered grid, and Gauss point sets for a 2D quadrilateral element.

\begin{figure}
\centering
\subfloat[GLL nodes]{\includegraphics[width=.3\textwidth]{figs/sbp.png}}
\hspace{.5em}
\subfloat[Staggered grid nodes]{\includegraphics[width=.3\textwidth]{figs/staggered.png}}
\hspace{.5em}
\subfloat[Gauss nodes]{\includegraphics[width=.3\textwidth]{figs/gsbp.png}}
\caption{Examples of nodal sets under which efficient entropy stable schemes can be constructed.  This work focuses on the construction of efficient and accurate SBP-SAT terms for Gauss nodal sets.}
\label{fig:nodesets}
\end{figure}

@@

Let $\bm{D}$ denote the nodal differentiation matrix, and let $\bm{V}_f$ denote the $(N+1)\times 2$ matrix which interpolates polynomials at Gauss nodes to values at endpoints.  These two matrices are defined entrywise as
\[
\bm{D}_{ij} = \LRl{\pd{\ell_j}{x}}_{x = x_i}, \qquad \LRp{\bm{V}_f}_{i1} = \ell_i(-1), \qquad  \LRp{\bm{V}_f}_{i2} = \ell_i(1).
\]
We also introduce the diagonal matrix of quadrature weights $\bm{W}_{ij} = \delta_{ij} w_i$.  In this work, $\bm{W}$ is identical to the mass matrix, whose entries are evaluated as $L^2$ inner products of basis functions.  This equality holds so long as the basis functions are Lagrange polynomials and integrals are approximated using quadrature (either GLL or Gauss).  
%\[
%\bm{M}_{ij} = \int_{-1}^{1} \ell_j(x)\ell_i(x) = \sum_{k=1}^{N+1}  w_k\ell_j(x_k)\ell_i(x_k)  = \delta_{ij} w_i.  
%\]
It can be show that the mass and differentiation matrices for Gauss nodes fall under the class of generalized SBP (GSBP) operators \cite{fernandez2014generalized}.  
\begin{lemma}
\label{lemma:sbp}
$\bm{Q} = \bm{W}\bm{D}$ satisfies the generalized summation by parts property
\[
\bm{Q} = \bm{V}_f^T \bm{B} \bm{V}_f - \bm{Q}^T, \qquad \bm{B} = \begin{bmatrix}-1 & \\ & 1\end{bmatrix}.
\]
\end{lemma}
%\begin{proof}
The proof is a direct restatement of integration by parts, and can be found in \cite{ortleb2016kinetic, ortleb2017kinetic, ranocha2018generalised}.  
%We reproduce it here for completeness.  Let $\bm{u},\bm{v}$ be two arbitrary vectors, and let $u(x), v(x)$ denote the polynomials whose nodal values are given by $\bm{u},\bm{v}$.  Then, 
%\begin{align*}
%\bm{v}^T\bm{Q}\bm{u} &= \bm{v}^T\bm{W}\bm{D}\bm{u} = \int_{-1}^1 \pd{u}{x}v = \LRl{uv}_{-1}^1 - \int_{-1}^1 u\pd{v}{x} \\
%&= \LRp{\bm{V}_f \bm{v}}^T\bm{B}\bm{V}_f\bm{u} - \LRp{\bm{D}\bm{v}}^T\bm{W}\bm{u} = \bm{v}^T\bm{V}_f^T\bm{B}\bm{V}_f \bm{u} - \bm{v}^T\bm{Q}^T\bm{u},
%\end{align*}
%where we have used that $\bm{W}$ is diagonal and that $(N+1)$-point Gauss quadrature is exact for the above integrands.  Taking $\bm{v} = \bm{e}_i$ and $\bm{u} = \bm{e}_j$ for $i,j = 1,\ldots, N+1$ (where $\bm{e}_i$ denotes the $i$th canonical vector) completes the proof.  
%\end{proof}
Lemma~\ref{lemma:sbp} holds for both GLL and Gauss nodes, and switching between these two nodal sets simply results in a redefinition of the matrices $\bm{D}, \bm{V}_f$.  For example, because GLL nodes include boundary points, the interpolation matrix $\bm{V}_f$ reduces to a generalized permutation matrix which extracts the nodal values associated with the left and right endpoints.  %Lemma~\ref{lemma:sbp} can be extended directly to higher dimensions through a tensor product construction.

@@

\subsection{Existing entropy stable SBP-SATs for generalized SBP operators}
\label{sec:gsbpsat}

@@

The construction of entropy conservative schemes relies on appropriate SBP-SATs for Gauss collocation schemes \cite{fernandez2014review, crean2017high, fernandez2018simultaneous}.  Let the rows of $\bm{V}_f$ be denoted as $\bm{t}_L, \bm{t}_R$ 
\[
\bm{V}_f = \begin{bmatrix}
\horzbar & \bm{t}_L & \horzbar\\
\horzbar & \bm{t}_R & \horzbar
\end{bmatrix}, \qquad \LRp{\bm{t}_L}_j = \ell_j(-1), \qquad \LRp{\bm{t}_R}_j = \ell_j(1).
\]
The inter-element coupling terms in \cite{fernandez2014review, crean2017high, fernandez2018simultaneous} utilize a decomposition of the surface mass matrix $\bm{V}_f^T\bm{B}\bm{V}_f$ as  
\begin{equation}
\bm{V}_f^T\bm{B}\bm{V}_f 
= \bm{t}_R\bm{t}_R^T - \bm{t}_L\bm{t}_L^T.
%, \qquad \bm{V}_f = \begin{bmatrix}
%\bm{t}_L & \bm{t}_R
%\end{bmatrix}
\label{eq:bmatdecomp}
\end{equation}
The construction of entropy conservative schemes on multiple elements requires appropriate inter-element coupling terms (SBP-SATs) involving $\bm{t}_L, \bm{t}_R$.  We consider a two element mesh, and show how SBP-SATs can be interpreted as constructing a global SBP operator.

Let $\bm{u}^1_N, \bm{u}^2_N$ denote nodal degrees of freedom of the vector valued solution $\bm{u}(x)$ on the first and second element, respectively.  To simplify notation, we assume that all following operators are defined in terms of Kronecker products, such that they are applied to each component of $\bm{u}^1_N, \bm{u}^2_N$.  
We first define the matrix 
\begin{equation}
\bm{S} = \bm{Q} - \frac{1}{2}\bm{V}_f^T\bm{B}\bm{V}_f.
\label{eq:Smat}
\end{equation}
It is straightforward to show (using Lemma~\ref{lemma:sbp}) that $\bm{S}$ is skew-symmetric.  We can now define an SBP operator $\bm{D}_h = \bm{W}_h^{-1}\bm{Q}_h$ over two elements
\begin{equation}
\bm{Q}_{h} = 
\underbrace{\begin{bmatrix}
\bm{S}  & \frac{1}{2}\bm{t}_R\bm{t}_L^T \\
- \frac{1}{2}\bm{t}_L\bm{t}_R^T & \bm{S}
\end{bmatrix}}_{\bm{S}_h}
+
\underbrace{\begin{bmatrix}
-\frac{1}{2}\bm{t}_L\bm{t}_L^T  & \\
 & \frac{1}{2} \bm{t}_R\bm{t}_R^T
\end{bmatrix}}_{\frac{1}{2}\bm{B}_h}
, \qquad 
\bm{W}_h = \begin{bmatrix}\bm{W} &\\
& \bm{W}
\end{bmatrix}.
%\bm{D}_h = \begin{bmatrix}\bm{W}^{-1} &\\
%& \bm{W}^{-1}
%\end{bmatrix}
% \bm{Q}_h.
\label{eq:gsbp}
\end{equation}
It can be shown that $\bm{D}_h$ is high order accurate such that, if $\bm{u}_h$ is a polynomial of degree $N$, it is differentiated exactly.  Straightforward computations show that $\bm{Q}_h$ also satisfies an SBP property $\bm{Q}_h + \bm{Q}_h^T = \bm{B}_h$.
%\[
%\bm{Q}_h + \bm{Q}_h^T = \bm{B}_h.
%\begin{bmatrix}
%-\bm{t}_L\bm{t}_L  & \\
% & \bm{t}_R\bm{t}_R
%\end{bmatrix}.
%\]

>>>>

\newcommand{\Oh}{{\Omega_h}}
\renewcommand{\L}{L^2\LRp{\Omega}}
\newcommand{\LK}{L^2\LRp{\Omega^k}}
\newcommand{\LdK}{L^2\LRp{\partial \Omega^k}}
\newcommand{\Dhat}{\widehat{D}}
\newcommand{\Lhat}{L^2\LRp{\Dhat}}

@@

\newcommand{\LinfDk}{L^{\infty}\LRp{\Omega^k}}
\newcommand{\diag}[1]{{\rm diag}\LRp{#1}}
\newcommand{\Ksub}{K_{\rm sub}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\newcommand\myeq{\mathrel{\stackrel{\makebox[0pt]{\mbox{\normalfont\tiny def}}}{=}}}
%\newcommand\myeq{{=}}

@@

\begin{abstract}
The construction of high order entropy stable collocation schemes on quadrilateral and hexahedral elements has relied on the use of Gauss-Legendre-Lobatto collocation points \cite{fisher2013high, carpenter2014entropy, gassner2016split} and their equivalence with summation-by-parts (SBP) finite difference operators \cite{gassner2013skew}.  In this work, we show how to efficiently generalize the construction of semi-discretely entropy stable schemes on tensor product elements to Gauss points and generalized SBP operators.  Numerical experiments suggest that the use of Gauss points significantly improves accuracy on curved meshes.  
\end{abstract}

@@

%Time dependent nonlinear conservation laws are ubiquitous in computational fluid dynamics (CFD).  These conservation laws include the compressible Euler and Navier-Stokes equations, which govern compressible fluid behavior within aerospace applications, as well as the equations governing coastal flooding and geophysical flows (shallow water), and those describing astrophysical phenomena (magnetohydrodynamics).  The utility of these simulations depends both on the reliability and accuracy of numerical methods for the solution of the associated nonlinear partial differential equations (PDEs).  
Time dependent nonlinear conservation laws are ubiquitous in computational fluid dynamics, for which high order methods are increasingly of interest.  Such methods are more accurate per degree of freedom than low order methods, while also possessing much smaller numerical dispersion and dissipation errors.  This makes high order methods especially well suited to time-dependent simulations.  In this work, we focus specifically on discontinuous Galerkin methods on unstructured quadrilateral and hexahedral meshes.  These methods combine properties of high order approximations with the geometric flexibility of unstructured meshing.  %When paired with explicit time-stepping methods, high order DG and finite element methods are also well-suited for many-core and GPU computing \cite{klockner2009nodal, swirydowicz2017acceleration, wintermeyer2018entropy}.

@@

The instability of high order methods is rooted in the fact that discretizations of nonlinear conservation laws do not typically satisfy a discrete analogue of the conservation or dissipation of energy (entropy).  For low order methods, the lack of discrete stability can be offset by the presence of numerical dissipation, which serves as a stabilization mechanism.  Because high order methods possess low numerical dissipation, the absence of a discrete stability property becomes more noticeable, manifesting itself through increased sensitivity and instability.

The dissipation of entropy serves as an energetic principle for nonlinear conservation laws \cite{dafermos2005compensated}, and requires the use of the chain rule in its proof.  Discrete instability is typically tied to the fact that, when discretizing systems of nonlinear PDEs, the chain rule does not typically hold at the discrete level.  The lack of a chain rule was circumvented by using a non-standard ``flux differencing'' formulation \cite{fisher2013high, carpenter2014entropy, gassner2016split, gassner2017br1}, which is key to constructing semi-discretely entropy stable high order schemes on unstructured quadrilateral and hexahedral meshes.  Flux differencing replaces the derivative of the nonlinear flux with the discrete differentiation of an auxiliary quantity.  This auxiliary quantity is computed through the evaluation of a two-point entropy conservative flux \cite{tadmor1987numerical} using pairs of solution values at quadrature points.  These entropy stable schemes were later extended to non-tensor product elements using GLL-like quadrature points on triangles and tetrahedra \cite{chen2017entropy, crean2018entropy}.  More recently, the construction of efficient entropy stable schemes was extended to more arbitrary choices of basis and quadrature \cite{chan2017discretely, chan2018discretely}.

@@

The main motivation for exploring tensor product (quadrilateral and hexahedral) elements is the significant reduction in the number of operations required compared to high order entropy stable schemes on simplicial meshes \cite{chan2017discretely, chan2018discretely}.  Entropy stability schemes on simplicial elements require evaluating two-point fluxes between solution states at all quadrature points on an element.  For a degree $N$ approximation, the number of quadrature points on a simplex scales as $O(N^d)$ in $d$ dimensions, and results in $O(N^{2d})$ two-point flux evaluations per element.  In contrast, entropy stable schemes on quadrilateral and hexahedral elements require only the evaluation of two-point fluxes along lines of nodes due to a tensor product structure, resulting in $O(N^{d+1})$ evaluations in $d$ dimensions.

@@

We are interested in methods for the numerical solution of systems of conservation laws in $d$ dimensions
\begin{equation}
\pd{\bm{u}}{t} + \sum_{i=1}^d \pd{\bm{f}_i\LRp{\bm{u}}}{x_i} = 0,
\label{eq:nonlinpde}
\end{equation}
where $\bm{u}$ denotes the conservative variables, $\bm{f}_i(\bm{u})$ are nonlinear fluxes, and $x_i$ denotes the $i$th coordinate.  Many physically motivated conservation laws admit a statement of stability involving a convex scalar entropy $S(\bm{u})$.  We first define the entropy variables $\bm{v}(\bm{u})$ to be the gradient of the entropy $S(\bm{u})$ with respect to the conservative variables 
\[
\bm{v} \myeq \pd{S(\bm{u})}{\bm{u}}.  
\]
For a convex entropy, $\bm{v}(\bm{u})$ defines an invertible mapping from conservative to entropy variables.  We denote the inverse of this mapping (from entropy to conservative variables) by $\bm{u}(\bm{v})$.

At the continuous level, it can be shown (for example, in \cite{dafermos2005compensated}) that vanishing viscosity solutions to (\ref{eq:nonlinpde}) satisfy the strong form of an entropy inequality 
\begin{gather}
\pd{S(\bm{u})}{t} + \sum_{i=1}^d \pd{F_i(\bm{u})}{x_i} \leq 0, \qquad F_i(\bm{u}) = \bm{v}^T\pd{\bm{f}_i}{\bm{u}}, 
\label{eq:entropyineq}
\end{gather}
where $F_i$ denotes the $i$th scalar entropy flux function.  Integrating (\ref{eq:entropyineq}) over a domain $\Omega$ and applying the divergence theorem yields an integrated entropy inequality
\begin{equation}
\int_{\Omega} \pd{S(\bm{u})}{t} + \int_{\partial \Omega} \sum_{i=1}^d \bm{n}_i \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})} \leq 0,
\label{eq:weakentropyineq}
\end{equation}
where $\psi_i(\bm{u}) = \bm{v}^T\bm{f}_i(\bm{u}) - F_i(\bm{u})$ denotes the $i$th entropy potential, $\partial \Omega$ denotes the boundary of $\Omega$ and $\bm{n}_i$ denotes the $i$th component of the outward normal on $\partial \Omega$.  Roughly speaking, this implies that the time rate of change of entropy is less than or equal to the flux of entropy through the boundary.

@@

This comparison is motivated by the accuracy of each respective quadrature rule.  While $(N+1)$-point GLL quadrature rules are exact for polynomial integrands of degree $(2N-1)$, $(N+1)$-point Gauss quadrature rules are exact for polynomials of degree $(2N+1)$.  This higher accuracy of Gauss quadrature has been shown to translate to lower errors and slightly improved rates of convergence in simulations of wave propagation and fluid flow \cite{kopriva2010quadrature, hindenlang2012explicit, chan2015gpu}.  However, Gauss points have not been widely used to construct entropy stable discretizations due to the lack of efficient, stable, and high order accurate inter-element coupling terms, known as simultaneous approximation terms (SAT) in the finite difference literature \cite{fernandez2014review, crean2017high, fernandez2018simultaneous}.  SATs for Gauss points are non-compact, in the sense that they introduce all-to-all coupling between degrees of freedom on neighboring elements.  This results in greater communication between elements, as well as a significantly larger number of two-point flux evaluations and floating point operations.

It is possible to realize the improved accuracy of Gauss points while avoiding non-compact SATs through a staggered grid formulation, where the solution is stored at Gauss nodes but interpolated to a set of higher degree $(N+2)$ GLL ``flux'' points for computation \cite{parsani2016entropy}.  Because GLL nodes include boundary points, compact and high order accurate SAT terms can be constructed for the flux points.  After performing computations on the flux points, the results are interpolated back to Gauss points and used to evolve the solution forward in time.  Figure~\ref{fig:nodesets} shows an illustration of GLL, staggered grid, and Gauss point sets for a 2D quadrilateral element.

\begin{figure}
\centering
\subfloat[GLL nodes]{\includegraphics[width=.3\textwidth]{figs/sbp.png}}
\hspace{.5em}
\subfloat[Staggered grid nodes]{\includegraphics[width=.3\textwidth]{figs/staggered.png}}
\hspace{.5em}
\subfloat[Gauss nodes]{\includegraphics[width=.3\textwidth]{figs/gsbp.png}}
\caption{Examples of nodal sets under which efficient entropy stable schemes can be constructed.  This work focuses on the construction of efficient and accurate SAT terms for Gauss nodal sets.}
\label{fig:nodesets}
\end{figure}

@@

Let $\bm{D}$ denote the nodal differentiation matrix, and let $\bm{V}_f$ denote the $(N+1)\times 2$ matrix which interpolates polynomials at Gauss nodes to values at endpoints.  These two matrices are defined entrywise as
\[
\bm{D}_{ij} = \LRl{\pd{\ell_j}{x}}_{x = x_i}, \qquad \LRp{\bm{V}_f}_{1i} = \ell_i(-1), \qquad  \LRp{\bm{V}_f}_{2i} = \ell_i(1).
\]
We also introduce the diagonal matrix of quadrature weights $\bm{W}_{ij} = \delta_{ij} w_i$, as well as the one-dimensional mass matrix $\bm{M}$ whose entries are $L^2$ inner products of basis functions.  We assume that these inner products are computed using the quadrature rule $(x_i, w_i)$ at which the solution is collocated.  Under such an assumption, the mass matrix is diagonal with entries equal to the quadrature weights
\[
\bm{M}_{ij} = \int_{-1}^1 \ell_i(x)\ell_j(x) \approx \sum_{k=1}^{N+1} \ell_i(x_k)\ell_j(x_k) w_k = \delta_{ij} w_i = \bm{W}_{ij}.
\]
Since $\bm{M} = \bm{W}$ under a collocation assumption, we utilize $\bm{W}$ for the remainder of this work to emphasize that the mass matrix is diagonal and related to the quadrature weights $w_i$.  The treatment of non-diagonal mass matrices is covered in \cite{chan2017discretely, chan2018discretely}.

%In this work, we assume that $\bm{W}$ is identical to the mass matrix, which \note{fix and mention mass lumping} whose entries are evaluated as $L^2$ inner products of basis functions.  This equality holds so long as the basis functions are Lagrange polynomials and integrals are approximated using quadrature (either GLL or Gauss).  
%\[
%\bm{M}_{ij} = \int_{-1}^{1} \ell_j(x)\ell_i(x) = \sum_{k=1}^{N+1}  w_k\ell_j(x_k)\ell_i(x_k)  = \delta_{ij} w_i.  
%\]
It can be show that the mass and differentiation matrices for Gauss nodes fall under the class of generalized SBP (GSBP) operators \cite{fernandez2014generalized}.  
\begin{lemma}
\label{lemma:sbp}
$\bm{Q} = \bm{W}\bm{D}$ satisfies the generalized summation by parts property
\[
\bm{Q} = \bm{V}_f^T \bm{B} \bm{V}_f - \bm{Q}^T, \qquad \bm{B} = \begin{bmatrix}-1 & \\ & 1\end{bmatrix}.
\]
\end{lemma}
%\begin{proof}
The proof is a direct restatement of integration by parts, and can be found in \cite{fernandez2014generalized, ortleb2016kinetic, ortleb2017kinetic, ranocha2018generalised}.  
%We reproduce it here for completeness.  Let $\bm{u},\bm{v}$ be two arbitrary vectors, and let $u(x), v(x)$ denote the polynomials whose nodal values are given by $\bm{u},\bm{v}$.  Then, 
%\begin{align*}
%\bm{v}^T\bm{Q}\bm{u} &= \bm{v}^T\bm{W}\bm{D}\bm{u} = \int_{-1}^1 \pd{u}{x}v = \LRl{uv}_{-1}^1 - \int_{-1}^1 u\pd{v}{x} \\
%&= \LRp{\bm{V}_f \bm{v}}^T\bm{B}\bm{V}_f\bm{u} - \LRp{\bm{D}\bm{v}}^T\bm{W}\bm{u} = \bm{v}^T\bm{V}_f^T\bm{B}\bm{V}_f \bm{u} - \bm{v}^T\bm{Q}^T\bm{u},
%\end{align*}
%where we have used that $\bm{W}$ is diagonal and that $(N+1)$-point Gauss quadrature is exact for the above integrands.  Taking $\bm{v} = \bm{e}_i$ and $\bm{u} = \bm{e}_j$ for $i,j = 1,\ldots, N+1$ (where $\bm{e}_i$ denotes the $i$th canonical vector) completes the proof.  
%\end{proof}
Lemma~\ref{lemma:sbp} holds for both GLL and Gauss nodes, and switching between these two nodal sets simply results in a redefinition of the matrices $\bm{D}, \bm{V}_f$.  For example, because GLL nodes include boundary points, the interpolation matrix $\bm{V}_f$ reduces to a generalized permutation matrix which extracts the nodal values associated with the left and right endpoints.  %Lemma~\ref{lemma:sbp} can be extended directly to higher dimensions through a tensor product construction.

@@

\subsection{Existing entropy stable SATs for generalized SBP operators}
\label{sec:gsbpsat}

@@

The construction of entropy conservative schemes relies on appropriate SATs for Gauss collocation schemes \cite{fernandez2014review, crean2017high, fernandez2018simultaneous}.  Let the rows of $\bm{V}_f$ be denoted by column vectors $\bm{t}_L, \bm{t}_R$ 
\[
\bm{V}_f = \begin{bmatrix}
%\horzbar & \bm{t}_L^T & \horzbar\\
%\horzbar & \bm{t}_R^T & \horzbar
\LRp{\bm{t}_L}_1, & \ldots, & \LRp{\bm{t}_L}_{N+1}\\
\LRp{\bm{t}_R}_1, & \ldots, & \LRp{\bm{t}_R}_{N+1}
\end{bmatrix}, \qquad \LRp{\bm{t}_L}_j = \ell_j(-1), \qquad \LRp{\bm{t}_R}_j = \ell_j(1).
\]
The inter-element coupling terms in \cite{fernandez2014review, crean2017high, fernandez2018simultaneous} utilize a decomposition of the surface matrix $\bm{V}_f^T\bm{B}\bm{V}_f$ as  
\begin{equation}
\bm{V}_f^T\bm{B}\bm{V}_f 
= \bm{t}_R\bm{t}_R^T - \bm{t}_L\bm{t}_L^T.
%, \qquad \bm{V}_f = \begin{bmatrix}
%\bm{t}_L & \bm{t}_R
%\end{bmatrix}
\label{eq:bmatdecomp}
\end{equation}
The construction of entropy conservative schemes on multiple elements requires appropriate inter-element coupling terms (SATs) involving $\bm{t}_L, \bm{t}_R$.  We consider a two element mesh, and show how when coupled with SATs, the resulting discretization matrices can be interpreted as constructing a global SBP operator.

Let $\bm{u}^1_N, \bm{u}^2_N$ denote nodal degrees of freedom of the vector valued solution $\bm{u}(x)$ on the first and second element, respectively.  To simplify notation, we assume that all following operators are defined in terms of Kronecker products \cite{chen2017entropy}, such that they are applied to each component of $\bm{u}^1_N, \bm{u}^2_N$.  
We first define the matrix 
\begin{equation}
\bm{S} \myeq \bm{Q} - \frac{1}{2}\bm{V}_f^T\bm{B}\bm{V}_f.
\label{eq:Smat}
\end{equation}
It is straightforward to show (using Lemma~\ref{lemma:sbp}) that $\bm{S}$ is skew-symmetric.  We can now define an SBP operator $\bm{D}_h = \bm{W}_h^{-1}\bm{Q}_h$ over two elements
\begin{equation}
\bm{Q}_{h} \myeq 
\underbrace{\begin{bmatrix}
\bm{S}  & \frac{1}{2}\bm{t}_R\bm{t}_L^T \\
- \frac{1}{2}\bm{t}_L\bm{t}_R^T & \bm{S}
\end{bmatrix}}_{\bm{S}_h}
+
\underbrace{\begin{bmatrix}
-\frac{1}{2}\bm{t}_L\bm{t}_L^T  & \\
 & \frac{1}{2} \bm{t}_R\bm{t}_R^T
\end{bmatrix}}_{\frac{1}{2}\bm{B}_h}
, \qquad 
\bm{W}_h \myeq \begin{bmatrix}\bm{W} &\\
& \bm{W}
\end{bmatrix}.
%\bm{D}_h = \begin{bmatrix}\bm{W}^{-1} &\\
%& \bm{W}^{-1}
%\end{bmatrix}
% \bm{Q}_h.
\label{eq:gsbp}
\end{equation}
It can be shown that $\bm{D}_h$ is high order accurate such that, if $\bm{u}_h$ is a polynomial of degree $N$, it is differentiated exactly.  Straightforward computations show that $\bm{Q}_h$ also satisfies an SBP property $\bm{Q}_h + \bm{Q}_h^T = \bm{B}_h$.
%\[
%\bm{Q}_h + \bm{Q}_h^T = \bm{B}_h.
%\begin{bmatrix}
%-\bm{t}_L\bm{t}_L  & \\
% & \bm{t}_R\bm{t}_R
%\end{bmatrix}.
%\]

Ignoring boundary conditions, an entropy conservative scheme for (\ref{eq:nonlinpde}) on two elements can then be given as
\begin{gather}
\td{}{t}\bm{W}_h\bm{u}_h + 2\LRp{\bm{Q}_h \circ \bm{F}_S}\bm{1} = 0, \qquad \bm{u}_h = \begin{bmatrix}
\bm{u}^1_N\\
\bm{u}^2_N
\end{bmatrix} \label{eq:ESgsbp} \\
\LRp{\bm{F}_S}_{ij} = \bm{f}_S\LRp{\LRp{\bm{u}_h}_i,\LRp{\bm{u}_h}_j}, \qquad 1 \leq  i,j \leq 2(N+1), \nonumber
\end{gather}
where $\circ$ denotes the Hadamard product \cite{horn2012matrix}.  {It should be emphasized that here, $(\bm{u}_h)_i, (\bm{u}_h)_j$ denote vectors containing solution components at nodes $i,j$, and that (because $\bm{f}_S$ is a vector-valued flux) the term $\LRp{\bm{F}_S}_{ij}$ should be interpreted as a diagonal matrix whose diagonal entries consist of the components of $\bm{f}_S\LRp{(\bm{u}_h)_i, (\bm{u}_h)_j}$.}
